---
title: "Using and adapting the mizer extension template"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Using and adapting the mizer extension template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mizer)
library(mizerExtensionTemplate)
```

# What this package is

`mizerExtensionTemplate` is a working mizer extension package that you can
clone and adapt. Every line of code is commented to explain *what* is being
done and *why*. Read the source files alongside this vignette.

For the full conceptual background see:

- `vignette("extending-mizer", package = "mizer")` — all five extension
  mechanisms with worked examples.
- `vignette("creating-extension-packages", package = "mizer")` — turning
  a script into a composable, shareable package.

# The extension at a glance

`mizerExtensionTemplate` adds three things to a standard mizer model:

| Mechanism | What it adds | Where |
|-----------|-------------|-------|
| `setExtEncounter()` | Fixed allometric extra food | `constructor.R` |
| `setExtMort()` | Fixed background mortality | `constructor.R` |
| `projectEncounter` S3 method | Seasonal encounter multiplier | `rate-methods.R` |
| `setComponent("plankton")` | Dynamical plankton spectrum | `constructor.R` + `component-functions.R` |
| `getBiomass` S3 methods | Includes plankton in output | `generic-methods.R` |

# Quick start

```{r quick-start, eval = FALSE}
params <- newExtensionTemplateParams(NS_species_params)
sim    <- project(params, t_max = 10)
plotBiomass(sim)   # Plankton column appears automatically
```

```{r quick-start-run, echo = FALSE}
params <- newExtensionTemplateParams(NS_species_params)
sim    <- project(params, t_max = 10)
plotBiomass(sim)
```

# How the seasonal encounter works

The seasonal multiplier peaks at `t = 0.25` (quarter-year) and troughs at
`t = 0.75`. With the default amplitude of 0.2 the encounter rate varies by
±20 % around its annual mean.

```{r seasonal}
params_s <- newExtensionTemplateParams(NS_species_params, season_amplitude = 0.4)
enc_t0   <- getEncounter(params_s, t = 0)    # multiplier = 1.0
enc_t025 <- getEncounter(params_s, t = 0.25) # multiplier = 1.4
range(enc_t025 / enc_t0, na.rm = TRUE)
```

This seasonal effect is implemented in
`projectEncounter.mizerExtensionTemplate()` (`rate-methods.R`). It calls
`NextMethod()` to get the base encounter rate (including the plankton
contribution), then multiplies by the seasonal factor. Using a `project*`
method rather than `setRateFunction()` means another extension package can
also modify the encounter rate without conflict.

# How the plankton component works

The plankton is a vector spectrum stored on the full resource size grid.
At each time step:

1. `planktonEncounter()` adds the plankton encounter contribution to
   `getEncounter()` by presenting the plankton as an extra resource.
2. `planktonDynamics()` updates the plankton by solving the ODE
   `dP/dt = r(K − P) − m·P` analytically over the time step, where `m` is
   the predation mortality on the plankton spectrum.

```{r plankton-diet}
plotDiet(params, species = "Cod")
```

# Adapting this template for your extension

1. **Rename the package**: search and replace `mizerExtensionTemplate` → your
   package name throughout all files, including `DESCRIPTION`, the R files,
   `NAMESPACE`, and this vignette.

2. **Decide: metadata-only or dispatching?**
   - *Metadata-only* (like `mizerStarvation`): delete
     `mizerExtensionTemplate-class.R`, remove the `setClass()` calls, and
     remove the `coerceToExtensionClass()` call at the end of the
     constructor. Keep `params@extensions <- getRegisteredExtensions()`.
   - *Dispatching* (like `mizerShelf`): keep everything and define S3
     methods for the generics you need to override.

3. **Remove mechanisms you don't need**: each of the five mechanisms in
   `constructor.R` is independent. Delete the blocks that do not apply to
   your extension.

4. **Replace the plankton component** with your own component, or remove
   `setComponent()` entirely if you don't need a dynamical state variable.

5. **Run `devtools::document()`** to regenerate `NAMESPACE` from the roxygen2
   tags, then `devtools::test()` and `devtools::check()` before publishing.

# Checklist for dispatching extension authors

- [ ] `setClass("<myExt>", contains = "MizerParams")` and
  `setClass("<myExt>Sim", contains = "MizerSim")` in the class file.
- [ ] `mizer::registerExtension(pkgname, requirement = ...)` in `.onLoad`.
- [ ] Constructor ends with `params@extensions <- getRegisteredExtensions()`
  and `coerceToExtensionClass(params)`.
- [ ] Every S3 method is registered via `@method` + `@export`.
- [ ] Every S3 method calls `NextMethod()`.
- [ ] Rate modifications use `project*` methods, not `setRateFunction()`.
- [ ] Extension-specific state lives in `other_params(params)` or component
  params — not in new S4 slots.
