---
title: "Using and adapting the mizer extension template"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Using and adapting the mizer extension template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mizer)
library(mizerExtensionTemplate)
```

# What this package is

`mizerExtensionTemplate` is a working mizer extension package that you can
clone and adapt. Every line of code is commented to explain *what* is being
done and *why*. Read the source files alongside this vignette.

For the full conceptual background see:

- `vignette("guide-extend-mizer", package = "mizer")` — all five extension
  mechanisms with worked examples.
- `vignette("guide-create-extension-package", package = "mizer")` — turning
  a script into a composable, shareable package.

# The extension at a glance

`mizerExtensionTemplate` adds three things to a standard mizer model:

| Mechanism | What it adds | Where |
|-----------|-------------|-------|
| `setExtEncounter()` | Fixed allometric extra food | `constructor.R` |
| `setExtMort()` | Fixed background mortality | `constructor.R` |
| `projectEncounter` S3 method | Seasonal encounter multiplier | `rate-methods.R` |
| `setComponent("plankton")` | Dynamical plankton spectrum | `constructor.R` + `component-functions.R` |
| `getBiomass` S3 methods | Includes plankton in output | `generic-methods.R` |
| Classed arrays with a `type` | `planktonLevel()` plots as a proportion | `component-functions.R` |
| `signal_info()` | Reports a choice, obeying `info_level` | `constructor.R` |
| Bundled data object | `example_params` ready to use | `data/`, `R/data.R` |

# Bundled example model

The package ships a ready-made `example_params` object — a three-species
(Sprat, Herring, Cod) model built with `newExtensionTemplateParams()`. It is
stored in `data/example_params.rda` and lazy-loaded by R. Standard R
serialisation preserves its complete S3 class vector and extension metadata,
so no `.onLoad` hook or active binding is needed:

```{r example-params}
class(example_params)   # mizerExtensionTemplate, not plain MizerParams
getBiomass(example_params)  # Plankton entry is present
```

You can use it directly without calling `newExtensionTemplateParams()`:

```{r example-project, eval = FALSE}
sim <- project(example_params, t_max = 5)
plotBiomass(sim)
```

# Quick start (build your own)

```{r quick-start, eval = FALSE}
params <- newExtensionTemplateParams(NS_species_params)
sim    <- project(params, t_max = 10)
plotBiomass(sim)   # Plankton column appears automatically
```

```{r quick-start-run, echo = FALSE}
params <- newExtensionTemplateParams(NS_species_params)
sim    <- project(params, t_max = 10)
plotBiomass(sim)
```

# How the seasonal encounter works

The seasonal multiplier peaks at `t = 0.25` (quarter-year) and troughs at
`t = 0.75`. With the default amplitude of 0.2 the encounter rate varies by
±20 % around its annual mean.

```{r seasonal}
params_s <- newExtensionTemplateParams(NS_species_params, season_amplitude = 0.4)
enc_t0   <- getEncounter(params_s, t = 0)    # multiplier = 1.0
enc_t025 <- getEncounter(params_s, t = 0.25) # multiplier = 1.4
range(enc_t025 / enc_t0, na.rm = TRUE)
```

This seasonal effect is implemented in
`projectEncounter.mizerExtensionTemplate()` (`rate-methods.R`). It calls
`NextMethod()` to get the base encounter rate (including the plankton
contribution), then multiplies by the seasonal factor. Using a `project*`
method rather than `setRateFunction()` means another extension package can
also modify the encounter rate without conflict.

# How the plankton component works

The plankton is a vector spectrum stored on the full resource size grid.
At each time step:

1. `planktonEncounter()` adds the plankton encounter contribution to
   `getEncounter()` by presenting the plankton as an extra resource.
2. `planktonDynamics()` updates the plankton by solving the ODE
   `dP/dt = r(K − P) − m·P` analytically over the time step, where `m` is
   the predation mortality on the plankton spectrum.

```{r plankton-diet}
plotDiet(params, species = "Cod")
```

# Returning arrays that plot themselves properly

When your extension returns a size- or time-resolved quantity, wrap it in one of
mizer's array classes and say what kind of quantity the values are, with the
`type` argument. There are three types: `"value"` for a rate or an amount,
`"density"` for an amount per gram of body weight, and `"proportion"` for a
fraction.

`type` is not decoration — it decides how the array is plotted. A `"density"`
gets multiplied by the `dw/dl` Jacobian when it is plotted against a length
axis, because a density per gram is not a density per centimetre. A
`"proportion"` gets a linear y axis showing the whole of the interval from 0 to
1, so the value can be read against the scale it belongs to.

`planktonLevel()` (`component-functions.R`) is the template's example. It
returns the plankton abundance as a fraction of its carrying capacity, mirroring
mizer's own `resource_level()`:

```{r plankton-level}
lev <- planktonLevel(params)
attr(lev, "type")
```

Because it declares itself a proportion, plotting it needs no further
instruction — the y axis runs from 0 to 1 rather than being fitted to the data:

```{r plankton-level-plot}
plot(planktonLevel(params))
```

Declare `type` for every array you return, including when it is the default
`"value"` — as `getBiomass.mizerExtensionTemplateSim()` does. If you omit it,
mizer falls back to guessing from `value_name` and `units`, which is there for
backwards compatibility and is easy to fall foul of.

# Telling the user what your extension decided

When your extension makes a choice on the user's behalf, report it through
mizer's own mechanism rather than with `message()` or `warning()`, because a plain
`message()` ignores `info_level`, is not collected with the other reports, and is
swallowed on the `species_params<-()` path.

The constructor takes an `info_level` argument, forwards it to
`newMultispeciesParams()`, wraps its body in `with_info_level()` and raises one
report of its own with `signal_info()`. The result is that the template's report
arrives in the same block as mizer's, and obeys the same switch. The constructor
defaults to `default_info_level()`, so it follows the global `mizer_info_level`
option. Ask explicitly for all reports when you need deterministic output:

```{r info-level}
params_loud <- newExtensionTemplateParams(NS_species_params, info_level = 3)
```

The last line is the template's own. `info_level` decides how much it takes to
silence a report: ours is level 3, so
`info_level = 1` keeps mizer's important reports and drops ours.

```{r info-level-1}
params_terse <- newExtensionTemplateParams(NS_species_params, info_level = 1)
```

Keep this pattern in your own constructor: default the argument to
`default_info_level()` and forward it explicitly. Hard-coding `info_level` in
the call to `newMultispeciesParams()` makes a user who passes their own collide
with it.

# Adapting this template for your extension

1. **Rename the package**: search and replace `mizerExtensionTemplate` → your
   package name throughout all files, including `DESCRIPTION`, the R files,
   `NAMESPACE`, and this vignette.

2. **Decide: metadata-only or dispatching?**
   - *Metadata-only* (like `mizerStarvation`): delete
     `mizerExtensionTemplate-class.R` and remove the
     `coerceToExtensionClass()` call at the end of the constructor. Keep
     the `recordExtension()` call.
   - *Dispatching* (like `mizerShelf`): keep everything and define S3
     methods for the generics you need to override.

3. **Remove mechanisms you don't need**: each of the five mechanisms in
   `constructor.R` is independent. Delete the blocks that do not apply to
   your extension.

4. **Replace the plankton component** with your own component, or remove
   `setComponent()` entirely if you don't need a dynamical state variable.

5. **Run `devtools::document()`** to regenerate `NAMESPACE` from the roxygen2
   tags, then `devtools::test()` and `devtools::check()` before publishing.

# Checklist for dispatching extension authors

- [ ] **No** `setClass()` or class declaration: the extension names are entries
  in ordinary S3 class vectors managed by `coerceToExtensionClass()`.
- [ ] Constructor ends with `recordExtension()` (including the package version
  and installation requirement) and `coerceToExtensionClass(params)`.
- [ ] Every bundled `MizerParams` or `MizerSim` object is created through the
  extension's setup function before it is saved in `data/`.
- [ ] Every S3 method is registered via `@method` + `@export`.
- [ ] Every S3 method calls `NextMethod()`.
- [ ] Rate modifications use `project*` methods, not `setRateFunction()`.
- [ ] Extension-specific state lives in `other_params(params)` or component
  params — not in new top-level list elements.
- [ ] Every array returned to the user is wrapped in a mizer array class with an
  explicit `type` (`"value"`, `"density"` or `"proportion"`).
- [ ] Anything you tell the user goes through `signal_info()` inside a
  `with_info_level()`, never a bare `message()` or `warning()`; entry points take
  `info_level = default_info_level()` and forward it.
